Modeling the Latent Space of Symbolic String Quartet Music

### 2021-06-08 papers/final/alex/paper.tex
regenerate notebooks from updated scripts

\begin{abstract}
  This paper presents the results of a project to apply a recurrent
  variational autoencoder to the task of modeling a latent space for
  generating pieces of multi-instrument symbolic classical music
  through interpolation. The code utilized for this research is available at:
  \url{https://github.com/jackPhn/CSS586_Project/}
\end{abstract}

\keywords{deep learning, recurrent neural networks, sequential models,
  music modeling, latent space modeling, generative modeling}

@@

The state of the art in music generation still has a long way to go
before it can consistently generate music scores or performances that
would be enjoyable and popular for humans to listen to, but for this
reason it is an area of significant opportunity, where a number of
recent research projects have shown promising progress.

Google's \href{https://magenta.tensorflow.org/}{Magenta} is an
umbrella project for music deep learning research and development of
software tools to expose these models for use by creative artists and
students.

MusicVAE, part of the Magenta project, is a variational Long
Short-Term Memory (LSTM) autoencoder for MIDI that incorporates a
novel hierarchical structure using a ``conductor'' recurrent layer in
its decoder model to better capture structure at multiple levels and
avoid the problem of ``posterior/mode collapse'' whereby a generative
model learns to ignore its latent codes and rely on autoregression to
generate output sequences \cite{roberts_hierarchical_2018}. This model
is trained on 16-bar paragraphs of music and is capable of generating
new melodies that blend two given melodies together via latent space
interpolation. Among the literature surveyed, MusicVAE's approach is
the most similar to this work; in comparison MusicVAE dedicates
significant effort to devising an architecture that allows the model
to generate very long sequences of notes, which is beneficial for
other generation tasks but not strictly necessary for
interpolation. Our model is significantly simpler and demonstrates
that interpolation works well even on a single bar of music at a time.

@@

While there may be significant opportunity in future approaches that
combining the two, contemporary research generally treats symbolic and
audio music modeling as separate problem domains and this work does
the same, choosing to focus on symbolic music.

@@

In order to accommodate polyphonic music, we each MIDI file is
converted into a modified pianoroll representation, an example of
which is visualized in Figure \ref{pianoroll}. While the standard
pianoroll represents each time-step as a multi-hot encoded vector of
note values, because most notes are not being played at most timesteps
there is an extreme class imbalance and sparsity in the matrix. This
is reported in \cite{kumar2019polyphonic} to cause problems in model
fitting and metrics calculation, so they introduce an alternative
``multi-stream'' representation of the pianoroll, which we closely,
(but not exactly) emulate in this work.

In our modified representation, each instrument track is a sequence of
integers from 0-129 representing the MIDI note pitch value being
played at the given time step, where values 0-127 are pitches, 128 is
a rest (silence) and 129 is sustain, indicating that the previously
played pitch is held continuously. MIDI data contains pressure values
that indicate how hard or soft each note is played; this information
is discarded in our preprocessing and all notes are represented with
constant pressure. Because chords (multiple notes being played
simultaneously by the same instrument) are relatively rare in string
quartets, the lowest note of each chord is taken and the rest
discarded, to reduce dimensionality. Because notes at the extreme low
and high end are rare and are out of range of common instruments, the
note pitch values are clipped to a narrower range and ordinal
encoded. For the 36 string quartets found in the MusicNet database,
this results in 66 distinct values, where 0-63 represent notes played
by the violins, viola and cello, 64 is rest and 65 is sustain. Figure
\ref{pitch_dist} depicts the distribution of these pitch values per
instrument. Our software package also includes functions to convert
this representation back into a Music21 Score object, which can then
be written to a MIDI file.

@@

Because songs are typically at least a few minutes long and of varying
length, it will not be feasible to train with entire songs as
examples, so we will crop songs into phrases of equal numbers of
beats to use as training data.

The result of this preprocessing is that each training example isbe a
2D matrix of shape (time steps x instrument tracks) and stacking the
training examples produces a 3D tensor. Information regarding which
track corresponds to which musical instrument is lost and therefore
must be maintained separately to reconstruct the MIDI
representation. For string quartets this instrument ensemble is always
\texttt{[40, 40, 41, 42]}, which is an array of MIDI instrument codes
representing two violins, one viola and one cello. Tempo information
is also lost from this representation, so the outputs are normalized
to the default tempo of 120 beats per minute.

As the model will produce one output per instrument track, it can
incorporate only a fixed selection of instrument parts, similar to how
MusicVAE models three-part (drum, bass and melody)
\cite{roberts_hierarchical_2018} and MuseGAN models five-part (drum,
bass, guitar, string, piano) arrangements. Therefore this model is
purpose-built for string quartets, which are the most common type of
arrangement found in the MusicNet dataset.

@@

The encoder model utilizes an embedding layer to encode the note pitch
integers as vectors; embedding vector length is a tunable
hyperparameter and we found the best results with it set to 8. Then
two LSTM layers, with a dropout layer in between, convert the
sequences into latent codes, which are modeled as a multidimensional
Gaussian distribution $z$, produced by the encoder output Lambda layer
(Figure \ref{decoder}).

The decoder model samples from the latent space $z$ with the
distribution parameterized by the $\mu$ and $log(\sigma)$ values
learned by the encoder, and then utilizes two LSTM layers with a
dropout layer between them to generate sequences from the latent
codes. Then the network splits into four parallel outputs, one per
instrument track, where a fully connected layer with softmax
activation is used to select the most probable note value for the
the current track at the each time step (Figure \ref{decoder}).

@@

\cite{oore_this_2018} suggests augmentation via pitch shifting each
training example up or down by up to six semitones in order to create
additional training examples and reduce overfitting, so we followed
that recommendation and used it to double the size of the training
dataset. As a result, we saw much lower validation losses (compare
Figure \ref{loss} and Figure \ref{loss_augment} and a small but
noticeable improvement in the fidelity of the song reconstructions,
suggesting that this method was effective in reducing overfitting.

@@

The interpolations are between the following 18 pairs of pieces:

@@

Figures \ref{source_ranges} and \ref{generated_ranges} depict the
distribution of pitch range (highest pitch - lowest pitch) among
samples. The distributions appear similar; source distribution has a
mean of 30.5 while the generated distribution has a mean of 26.3, (the
medians are similar to the means for both) so the generated
distributions have a slightly narrower pitch range.

@@

Future work in this domain could involve disentanglement of the latent
dimensions to enable music style transfer and semantic editing via
interpolation in specific directions within the latent space. The VAE
model could be adapted to other musical genres aside from string
quartets, to interpolate within or across genres. The model could also
be modified into a conditional VAE to allow conditioning on the genre,
composer/artist name, or some other defining attribute label, to
influence the generated results.

>>>>

\begin{abstract}
  This paper presents the results of a project to apply a recurrent
  variational autoencoder to the task of modeling a latent space for
  generating pieces of multi-instrument symbolic classical music
  through linear interpolation. The code utilized for this research is
  available at: \url{https://github.com/jackPhn/CSS586_Project/}
\end{abstract}

\keywords{deep learning, recurrent neural networks, sequential models,
  autoencoders, music modeling, latent space modeling, generative
  modeling}

@@

The state of the art in music generation still has a long way to go
before it can consistently generate music scores or performances that
would be broadly enjoyable for humans to listen to, but for this
reason it is an area of significant opportunity, where a number of
recent research projects have shown promising progress.

Google's \href{https://magenta.tensorflow.org/}{Magenta} is an
umbrella project for music deep learning research and development of
software tools to expose these models for use by creative artists and
students. MusicVAE, part of the Magenta project, is a variational Long
Short-Term Memory (LSTM) autoencoder for MIDI that incorporates a
novel hierarchical structure using a ``conductor'' recurrent layer in
its decoder model to better capture structure at multiple levels and
avoid the problem of ``posterior/mode collapse'' whereby a generative
model learns to ignore its latent codes and rely on autoregression to
generate output sequences \cite{roberts_hierarchical_2018}. This model
is trained on 16-bar paragraphs of music and is capable of generating
new melodies that blend two given melodies together via latent space
interpolation. Among the literature surveyed, MusicVAE's approach is
the most similar to this work; in comparison MusicVAE dedicates
significant effort to devising an architecture that allows the model
to generate very long sequences of notes, which is beneficial for
other generation tasks but not strictly necessary for
interpolation. Our model is significantly simpler and demonstrates
that interpolation works well even on a single bar of music at a time.

@@

While there may be significant opportunity in future approaches that
combine the two, contemporary research generally treats symbolic and
audio music modeling as separate problem domains and this work does
the same, choosing to focus on symbolic music.

@@

In order to accommodate polyphonic music, each MIDI file is converted
into a modified pianoroll representation, an example of which is
visualized in Figure \ref{pianoroll}. While the standard pianoroll
represents each time-step as a multi-hot encoded vector of note
values, because most notes are not being played at most timesteps
there is an extreme class imbalance and sparsity in the matrix. This
is reported in Kumar and Ravindran (2019) \cite{kumar2019polyphonic}
to cause problems in model fitting and metrics calculation, so they
introduce an alternative ``multi-stream'' representation of the
pianoroll, which we closely, (but not exactly) emulate in this work.

In our modified representation, each instrument track is a sequence of
integers from 0-129 representing the MIDI note pitch value being
played at the given time step, where values 0-127 are pitches, 128 is
a rest (silence) and 129 is sustain, indicating that the previously
played pitch is held continuously. MIDI data contains pressure values
that indicate how hard or soft each note is played; this information
is discarded in our preprocessing and all notes are represented with
constant pressure. Because chords (multiple notes being played
simultaneously by the same instrument) are relatively rare in string
quartets, the lowest note of each chord is taken and the rest
discarded, to reduce dimensionality. Because notes at the extreme low
and high end are rare and are out of range of common instruments, the
note pitch values are clipped to a narrower range and ordinal
encoded. For the 36 string quartets found in the MusicNet database,
this results in 66 distinct values, where 0-63 represent notes played
by the violins, viola and cello, 64 is rest and 65 is sustain. Figure
\ref{pitch_dist} depicts the distribution of MIDI pitch values per
instrument and the degree to which they overlap.

Our software package also includes functions to convert this
representation back into a Music21 Score object, which can then be
written to a MIDI file that can be synthesized for human listening.

@@

Because songs are typically at least a few minutes long and of varying
length, it is not feasible to train with entire songs as examples, so
we crop songs into phrases of equal numbers of beats to use as
training data. The result of this preprocessing is that each training
example is a 2D matrix of shape (time steps x instrument tracks) and
stacking the training examples produces a 3D tensor. Information
regarding which track corresponds to which musical instrument is lost
and therefore must be maintained separately to reconstruct the MIDI
representation. For string quartets this instrument ensemble is always
\texttt{[40, 40, 41, 42]}, which is an array of MIDI instrument codes
representing two violins, one viola and one cello. Tempo information
is also lost from this representation, which uses symbolic time steps
rather than wall clock time, so the outputs are standardized to the
default tempo of 120 beats per minute.

As the model will produce one output per instrument track, it can
incorporate only a fixed selection of instrument parts, similar to how
MusicVAE models three-part (drum, bass and melody)
\cite{roberts_hierarchical_2018} and MuseGAN models five-part (drum,
bass, guitar, string, piano) arrangements. Therefore the number of
tracks is a hyperparameter to the network architecture definition and
this model is trained for string quartets, which are the most common
type of arrangement found in the MusicNet dataset.

@@

The encoder model utilizes an embedding layer to encode the note pitch
integers as vectors; embedding vector length is a tunable
hyperparameter and we found the best results with it set to 8. Then
two LSTM layers, with a dropout layer in between, convert the
sequences into latent codes, which are modeled as a 512-dimensional
Gaussian distribution $z$, produced by the encoder output Lambda layer
(Figure \ref{decoder}).

The decoder model samples from the latent space $z$ with the
distribution parameterized by the $\mu$ and $log(\sigma)$ values
learned by the encoder, and then utilizes two LSTM layers with a
dropout layer between them to generate sequences from the latent
codes. Then the network splits into four parallel outputs, one per
instrument track, where a fully connected layer with softmax
activation is used to select the most probable note value for the
the current track at each time step (Figure \ref{decoder}).

@@

\cite{oore_this_2018} suggests augmentation via pitch shifting each
training example up or down by up to six semitones in order to create
additional training examples and reduce overfitting, so we followed
that recommendation and used it to double the size of the training
dataset. As a result, we saw much lower validation losses (compare
Figure \ref{loss} and Figure \ref{loss_augment}) and a small but
noticeable improvement in the fidelity of the song reconstructions,
suggesting that this method was effective in reducing overfitting.

@@

The interpolations are between the following 18 pairs of pieces
denoted by their filenames in the original MusicNet reference MIDI
archive, which can be downloaded from:
\url{https://homes.cs.washington.edu/~thickstn/media/musicnet_midis.tar.gz}

@@

Figures \ref{source_ranges} and \ref{generated_ranges} depict the
distribution of pitch range (highest pitch - lowest pitch) among
samples. The distributions appear similar; the source distribution has
a mean of 30.5 while the generated distribution has a mean of 26.3,
(the medians are similar to the means for both) so the generated
distributions have a slightly narrower pitch range.

@@

Future work in this domain could involve disentanglement of the latent
dimensions to enable music style transfer and semantic editing via
interpolation in specific directions within the latent space. The VAE
model could also be adapted to other musical genres aside from string
quartets, to interpolate within or across genres.
